---
title: "Fifty-five years of life expectancy"
author: "Arunark"
date: 2026-09-23
categories: [r, visualization]
---

Description: "How life expectancy changed by continent from 1952 to 2007, and how it tracks income.

Data: [gapminder R package](https://jennybc.github.io/gapminder/) by Jennifer Bryan, an excerpt of data from [Gapminder](https://www.gapminder.org/data/). The package is released under CC0 and the underlying Gapminder data under CC BY 4.0.

The data covers 142 countries every five years from 1952 to 2007. Two questions: which continents gained the most life expectancy over that period, and how closely does life expectancy track income?

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false
library(gapminder)
library(dplyr)
library(ggplot2)
library(knitr)
```

## A first look

```{r}
#| label: overview
gapminder |>
  summarise(
    countries = n_distinct(country),
    continents = n_distinct(continent),
    first_year = min(year),
    last_year = max(year)
  )
```

## Gains by continent

```{r}
#| label: tbl-lifeexp-change
#| tbl-cap: "Median life expectancy in 1952 and 2007 by continent, sorted by gain."
change <- gapminder |>
  group_by(continent) |>
  summarise(
    life_1952 = median(lifeExp[year == 1952]),
    life_2007 = median(lifeExp[year == 2007]),
    gain = life_2007 - life_1952
  ) |>
  arrange(desc(gain))

kable(change, digits = 1)
```

@tbl-lifeexp-change shows that `r change$continent[1]` gained the most, about `r round(change$gain[1], 1)` years in median life expectancy. Africa started lowest and was still lowest in 2007.

## Life expectancy and income in 2007

```{r}
#| label: fig-life-gdp
#| fig-cap: "Life expectancy against GDP per capita in 2007. Point size is population. The x axis is on a log scale."
gap_2007 <- filter(gapminder, year == 2007)

ggplot(gap_2007, aes(gdpPercap, lifeExp, colour = continent, size = pop)) +
  geom_point(alpha = 0.7) +
  scale_x_log10(labels = scales::label_dollar()) +
  scale_size_area(guide = "none") +
  labs(x = "GDP per capita (log scale)", y = "Life expectancy (years)", colour = "Continent") +
  theme_minimal()
```

```{r}
#| label: model
fit <- lm(lifeExp ~ log10(gdpPercap), data = gap_2007)
coef(fit)
summary(fit)$r.squared
```

## What I found

On a log scale the relationship in @fig-life-gdp is close to linear. The model says each tenfold increase in GDP per capita goes with about `r round(coef(fit)[2], 1)` more years of life expectancy, and income alone explains `r scales::percent(summary(fit)$r.squared, accuracy = 1)` of the variation across countries in 2007. The points furthest below the line are mostly African countries, which matches Africa having the lowest median life expectancy in both 1952 and 2007 (@tbl-lifeexp-change).