---
title: "R and Python in one post"
author: "Arunark"
date: 2026-09-25
categories: [r, python, reticulate]
engine: knitr
---

Description: "Passing data between R and Python with reticulate."

Data: [`mtcars`](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html), built into R. Henderson, H. V. and Velleman, P. F. (1981), from the 1974 *Motor Trend* US magazine.

This post passes data from R to Python, fits a line in Python, and passes the fitted coefficients back to R for plotting. It uses the knitr engine, which sends `{python}` chunks to reticulate in the same session.

## Setup (R)

```{r}
#| label: setup
#| message: false
library(reticulate)
library(ggplot2)

# Chunks run with the working directory set to this post's folder,
# so the project's .venv is two levels up.
use_virtualenv(normalizePath(file.path("..", "..", ".venv")), required = TRUE)
```

## Prepare the data in R

```{r}
#| label: r-data
cars_df <- mtcars[, c("mpg", "wt", "hp")]
head(cars_df)
```

## Fit the model in Python

`r.cars_df` reads the R data frame into Python as a pandas DataFrame.

```{python}
#| label: py-fit
import sys
from pathlib import Path
import numpy as np

print(Path(sys.executable).resolve().parts[-3:])

cars = r.cars_df
slope, intercept = np.polyfit(cars["wt"], cars["mpg"], 1)
print(f"mpg = {intercept:.2f} - {abs(slope):.2f} * wt")
```

## Plot in R with the Python result

`py$slope` and `py$intercept` read the Python values back into R.

```{r}
#| label: fig-mpg-wt
#| fig-cap: "Fuel economy against weight, with the least-squares line fitted in Python."
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  geom_abline(intercept = py$intercept, slope = py$slope, colour = "steelblue") +
  labs(x = "Weight (1000 lbs)", y = "Miles per gallon") +
  theme_minimal()
```

Each extra 1000 lbs of weight costs about `r round(abs(py$slope), 1)` miles per gallon.